---
title: "Demonstration of single cell data processing on the brain data"
author: "Yongjin Park"
date: "`r Sys.time()`"
---

```{r include=FALSE}
library("R.utils")
library("lemon")
source("Util-rmd.R")

fig.dir = 'Fig/example/brain_2018-05-03/'
dir.create(fig.dir, recursive = TRUE, showWarnings = FALSE)
knitr::opts_chunk$set(fig.path = fig.dir, fig.width = 4, fig.height = 3)
options(stringsAsFactors = FALSE)
```

We assume the following executable files are properly installed:

* `mmutil_filter_row`

* `mmutil_spectral_cluster_col`

```{r}
Sys.time()
```

## Preprocessing

Create a working directory:

```{bash}
[ -d example/brain_2018-05-03/ ] || mkdir -p example/brain_2018-05-03/
```

### Select top `5000` most variable features

```{bash}
if ! [ -f example/brain_2018-05-03/top.mtx.gz ]; then
    mmutil_filter_row 5000 \
    data/brain_2018-05-03/filtered_count_matrix.mtx.gz \
    data/brain_2018-05-03/features.tsv.gz \
    example/brain_2018-05-03/top \
    2> /dev/null
fi

# just copy the barcode file to be consistent
if ! [ -f example/brain_2018-05-03/top.barcodes.tsv.gz ]; then
    cp data/brain_2018-05-03/barcodes.tsv.gz example/brain_2018-05-03/top.barcodes.tsv.gz
fi
```

Show the full spectrum of the feature scores:

```{r fig_var_score}

out.dir = "example/brain_2018-05-03/"

var.scores = out.dir %&% "top.full_scores.gz" %>%
    read_tsv(col_names="var", col_types="d") %>%
    mutate(idx = 1:n())

.gg.plot(var.scores, aes(x=idx, y=log(1 + var))) +
    geom_vline(xintercept=5000, lty=2, col='red') +
    geom_line(size=2) +
    ylab("log(1 + variance)") +
    xlab("ranked features")
```

## Run clustering

Meta data to compare with the previous annotations:

```{r}
meta.file = "data/brain_2018-05-03/filtered_column_metadata.txt"
meta.tab = fread(meta.file)
```

Let's visualize clustering results with the following function:

```{r}
.show.scatter = function(.argmax, .embed) {

    argmax.tab = fread(.argmax) %>%
        (function(x) { colnames(x) = c("TAG", "k"); x })

    embedded.tab = fread(.embed) %>%
        (function(x) { colnames(x) = c("e1", "e2"); x })

    ## Show 2D embedding of those clusters of cells

    .K = max(argmax.tab$k)
    .lab = c("cluster" %&% 1:.K, "unassigned")

    argmax.tab = argmax.tab[, k := factor(k, c(1:.K, -1), .lab)]

    .dt = cbind(argmax.tab, embedded.tab) %>% na.omit()

    p1 =
        .gg.plot(.dt, aes(x=e1, y=e2, color=as.factor(k))) +
        geom_point(size=.5)

    print(p1)

    ## Remove outlier cells
    .dt2 = merge(.dt, meta.tab, by = "TAG")[k != "unassigned"]

    p2 =
        .gg.plot(.dt2, aes(x=e1, y=e2, color=broad.cell.type)) +
        geom_point(size=.5)

    print(p2)
}
```

### Gaussian Mixture Model after spectral transformation

```{bash}
if ! [ -f example/brain_2018-05-03/gmm.argmax.gz ]; then
    mmutil_spectral_cluster_col --mtx example/brain_2018-05-03/top.mtx.gz \
                                --method GMM --trunc 10 --verbose --rank 5 \
                                --col example/brain_2018-05-03/top.barcodes.tsv.gz \
                                --out example/brain_2018-05-03/gmm 2> /dev/null
fi
```

Read and show GMM estimation results:

```{r fig_scatter_gmm}
.show.scatter(out.dir %&% "/gmm.argmax.gz",
              out.dir %&% "/gmm.embedded.gz")
```

### Density-based spectral clustering

A strong distributional assumption of underlying data often result in
poor clustering results while density-based non-parametric clustering
methods robustly identify clusters of data points distributed in an
arbitrary fashion.

```{bash}
if ! [ -f example/brain_2018-05-03/dbscan_level_1.argmax.gz ]; then
    mmutil_spectral_cluster_col --mtx example/brain_2018-05-03/top.mtx.gz --rank 5 \
                                --method DBSCAN --knn 50 --prune_knn --min_size 50 --verbose \
                                --col example/brain_2018-05-03/top.barcodes.tsv.gz \
                                --out example/brain_2018-05-03/dbscan 2> /dev/null
fi
```

Read and show DBSCAN estimation at level 5:

```{r fig_scatter_dbscan_5}
.show.scatter(out.dir %&% "/dbscan_level_5.argmax.gz",
              out.dir %&% "/dbscan_level_5.embedded.gz")
```

Read and show DBSCAN estimation at level 1:

```{r fig_scatter_dbscan_1}
.show.scatter(out.dir %&% "/dbscan_level_1.argmax.gz",
              out.dir %&% "/dbscan_level_1.embedded.gz")
```

Read and show DBSCAN estimation at level 9:

```{r fig_scatter_dbscan_9}
.show.scatter(out.dir %&% "/dbscan_level_9.argmax.gz",
              out.dir %&% "/dbscan_level_9.embedded.gz")
```


```{r}
Sys.time()
```
