---
title: "Cell type annotation by supervised von Mises-Fisher clustering"
author: "Yongjin Park"
date: "`r Sys.time()`"
---

```{r include=FALSE}
library(tidyverse)
library(data.table)
library(patchwork)
library("R.utils")
library("lemon")
source("Util-rmd.R")

fig.dir = 'Fig/example/annotation/'
dir.create(fig.dir, recursive = TRUE, showWarnings = FALSE)
knitr::opts_chunk$set(fig.path = fig.dir, fig.width = 4, fig.height = 3)
knitr::opts_chunk$set(dev.args = list(bg = "transparent"))
options(stringsAsFactors = FALSE)
```

## Data preparation

We downloaded a list of PsychENCODE marker gene annotations (broad cell types). The file looks like this:

```{bash}
cat PsychENCODE.marker | head
```

```{r Fig_Marker, fig.width=6, fig.height=1.5}
marker.genes =
  fread("PsychENCODE.marker",header=FALSE) %>%
  (function(x) { colnames(x) = c("geneSymbol", "celltype"); x })

## just show these marker genes
.oo = list(rows = rev(c("Ex", "In", "Oligo", "OPC", "Microglia", "Astro", "Per", "Endo")))

.oo$cols =
    marker.genes %>% 
    mutate(row = celltype, col = geneSymbol, weight = 1) %>%
    col.order(.oo$rows)

.df = marker.genes %>%
    mutate(row = factor(celltype, .oo$rows)) %>% 
    mutate(col = factor(geneSymbol, .oo$cols))
    
.gg.plot(.df, aes(y=row, x=col)) +
    geom_tile() +
    xlab("genes") + ylab("cell types") + 
    theme(axis.ticks.x=element_blank()) +
    theme(axis.text.x=element_blank())
```

For each cell, we can distinguish the identity of cell types by cosine similarity between them. For instance, we can construct a 0/1 marker gene membership matrix as follows:

```{r}
M = .df %>%
  select(celltype, geneSymbol) %>%
  mutate(val=1) %>%
  spread(key="celltype", value="val", fill=0) %>%
  select(-geneSymbol) %>%
  as.matrix  

M %>% head
```

We can measure the cosine similarity across the columns by taking
dot-product after normalization.

```{r}
M = apply(M, 2, function(x) x/sqrt(sum(x^2)))
t(M) %*% M %>% round(2)
```

PsychENCODE's marker gene selection might be imperfect since some
"marker" genes are shared by different cell types. However, it has
enough discriminatory power.

Using this initial dictionary of marker genes, we can start analyzing
10x data, which consist of these three files:

* Sparse count matrix:

```{bash}
gzip -cd data/brain_2018-05-03/filtered_count_matrix.mtx.gz | head
```

* Cell barcode list:

```{bash}
gzip -cd data/brain_2018-05-03/barcodes.tsv.gz | head
```

* Gene/feature list (**these names should exist in the marker file**):

```{bash}
gzip -cd data/brain_2018-05-03/features.tsv.gz | head
```

## Annotate columns of 10x count matrix.

Let's store all the results here:

```{bash}
[ -d example/annotate_brain ] || mkdir -p example/annotate_brain
```

Running this takes some time...

```{bash}
[ -f example/annotate_brain/out.annot.gz ] || \
	../src/mmutil_annotate_col \
		--mtx data/brain_2018-05-03/filtered_count_matrix.mtx.gz \
		--feature data/brain_2018-05-03/features.tsv.gz \
		--balance_markers \
		--col data/brain_2018-05-03/barcodes.tsv.gz \
		--ann PsychENCODE.marker \
		--batch_size 1000 \
		--out example/annotate_brain/out 2> /dev/null
```

That is it! We have evaluated the probability of cell type assignment
for these cells:

```{bash}
gzip -cd example/annotate_brain/out.annot.gz | head
```

* The first column: barcode ID
* The second column: cell type provided by the `.marker` file
* The third column: the probability of most probable cell type

The algorithm iteratively adjust marker gene contribution:

```{r Fig_Marker_adj, fig.width=6, fig.height=1.5}
.labels = fread("example/annotate_brain/out.label_names.gz",header=FALSE) %>% unlist

.genes = fread("example/annotate_brain/out.marker_names.gz",header=FALSE) %>% unlist

.df = fread("example/annotate_brain/out.marker_profile.gz") %>%
  (function(x) { colnames(x) = .labels; x }) %>% 
  as_tibble() %>% 
  mutate(geneSymbol = .genes) %>%
  gather(key=celltype, value=phi, -geneSymbol) %>%
  group_by(geneSymbol, celltype) %>%
  slice(which.max(phi)) %>% 
  ungroup() %>% 
  mutate(row = factor(celltype, .oo$rows)) %>% 
  mutate(col = factor(geneSymbol, .oo$cols))

.gg.plot(.df, aes(y=row, x=col, fill = phi)) +
  theme(legend.position = "top") +
  geom_tile() +
  scale_fill_gradientn(colours=c("white", "red")) +
  xlab("genes") + ylab("cell types") + 
  theme(axis.ticks.x=element_blank()) +
  theme(axis.text.x=element_blank())

```

## Embedding to show the cell type annotation result

```{bash}
[ -f example/annotate_brain/spectral.sample_factor.gz ] || \
	../src/mmutil_spectral_col \
		--mtx data/brain_2018-05-03/filtered_count_matrix.mtx.gz \
		--rank 30 --em_iter 10 --em_tol 1e-4 --log_scale --block_size 1000 \
		--out example/annotate_brain/spectral 2> /dev/null
```


```{r fit_umap}
.file = "example/annotate_brain/umap.gz"
if(!file.exists(.file)) {
    V = fread("example/annotate_brain/spectral.sample_factor.gz") %>% as.matrix
    uu.dt = uwot::umap(V, n_neighbors = 50,  n_threads = 16, verbose = TRUE) %>%
        as.data.table
    .fwrite(uu.dt, file=.file)
} else {
    uu.dt = fread(.file)
}
```



```{r Fig_embedding_umap, fig.width=6, fig.height=5}
.pred = fread("example/annotate_brain/out.annot.gz",header=FALSE)
colnames(.pred) = c("TAG", "prediction", "probability", "log.llik")
uu.dt = cbind(uu.dt, .pred)

.gg.plot(uu.dt, aes(x = V1, y = V2, color = prediction)) +
    xlab("umap 1") + ylab("umap 2") +
    geom_point(alpha = .5, size = .5) +
    scale_color_brewer("prediction", type="div", palette="Paired")
```

